---
title: "Depression and Anxiety Rates related to Covid-19"
team members: "Alexa Elliott, Olivia Riniker, Connor Boysen"
date: "2024-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Depression and Anxiety rates related to Covid-19

### Alexa Elliot, Olivia Riniker, Connor Boysen

## Introduction

This data set can asses the social and economic impacts that Covid-19 had on people in America. Using the frequency of symptoms of anxiety and depression over a seven day period.To asses these different variables we will explore these questions:

1.  What is the prevalence of anxiety and depression symptoms among different age groups in the United States during the COVID-19 pandemic?

2.  How does the severity of anxiety and depression change across time w.r.t Education?

3.  Are there disparities in reported anxiety and depression symptoms based on factors such as age, gender, race/ethnicity, education level, or geographic location?

4.  What are the potential implications of the findings for public health interventions and mental wellness support during and after the pandemic?

5.  If a lock down were to ever happen again in what ways can we better address the emotional states of american citizens based on facts and statistics.

## Data Structure

the link for our data set is <https://data.cdc.gov/api/views/8pt5-q6wp/rows.csv?accessType=DOWNLOAD> . We obtained this data set from data.gov, it is a data set from the the U.S. Department of Health & Human Services which they obtained through an internet questionnaire.This data covers from April of 2020 to march or 2024 so we will be able to see the affects of the lock down as well as the change produced after the knockdown was disbanded.We are not include confidence interval(because we already have the seperate high and low), the quartile range(this column is missing many values), and time period label(because we have specific start and end dates.)

**pick variables to exclude that won't be helpful**

## cleaning data

First we load in the csv file for this data set.

```{r}
library(readr)
anxiety_depression <- read_csv("Indicators_of_Anxiety_or_Depression_Based_on_Reported_Frequency_of_Symptoms_During_Last_7_Days.csv")
View(anxiety_depression)

```

next we will start cleaning the data, starting with selecting the variables that we will be exploring in this data.

```{r}
# deleting unneeded columns
library(tidyverse)
col_delete <- c("Time Period Label", "Confidence Interval", "Quartile Range")

anxiety_depression <- anxiety_depression %>% 
  select(-one_of(col_delete))


#seperating the data by indicator
# Filter data for depressive disorder
depressive_data <- anxiety_depression %>%
  filter(Indicator == "Symptoms of Depressive Disorder")

# Filter data for anxiety disorder
anxiety_data <- anxiety_depression %>%
  filter(Indicator == "Symptoms of Anxiety Disorder")

# Filter data for depressive and anxiety disorder
depressive_anxiety_data <- anxiety_depression %>%
  filter(Indicator == "Symptoms of Anxiety Disorder or Depressive Disorder")

```

##Variables 
- Indicator: indicates the type of mental health condition being measures (e.g. symptoms of depressing disorder, anxiety disorder, or both)

- Group: Describes the subgrouping or category of the data (e.g., National Estimate, By Age, By Sex, By Race/Hispanic ethnicity, By Education, By State).

- State: Specifies the state or region within the United States for which the data is collected.

- Subgroup: Further specifies the subgroup within the specified group (e.g., age range, sex, race/ethnicity, education level, state).

- Phase: Represents the phase or wave of data collection (e.g., Phase 1, Phase 2).

- Time Period: Numeric representation of the time period during which the data was collected.

- Time Period Start Date: The start date of the time period.

- Time Period End Date: The end date of the time period.

- Value: The reported value or percentage representing the frequency of symptoms of the specified mental health condition.

- Low Cl: The lower bound of the confidence interval for the reported value.

- High Cl: The upper bound of the confidence interval for the reported value.

```{r}
# Load the required library
library(tidyverse)

by_ages <- depressive_anxiety_data %>%  
  filter(Group == "By Age")


ggplot(data = by_ages, aes(x = Subgroup, y = Value)) +
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Prevalence of Anxiety and Depression Symptoms Among Different Age Groups",
       x = "Age Groups",
       y = "Prevalence (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The bar plot illustrates the prevalence of anxiety and depression symptoms among various age groups in the United States during the COVID-19 pandemic. As depicted, there is a noticeable trend indicating that younger age groups tend to exhibit a higher prevalence of these symptoms compared to older age groups. Specifically, individuals in the 18-29 years age range demonstrate the highest prevalence, followed by those in the 30-39 years age bracket, and so forth. This suggests that younger adults are more susceptible to experiencing symptoms of anxiety and depressive disorders during the pandemic. Such findings may reflect the unique challenges faced by younger demographics, including disruptions to education, employment, and social interactions, all of which have been significantly impacted by the pandemic. These insights underscore the importance of targeted mental health interventions and support systems, particularly for younger populations, to mitigate the adverse psychological effects of the COVID-19 crisis.


Question #2: How does the severity of anxiety and depression change across time w.r.t Education?