---
title: "Water Quality.rmd"
team members: "Alexa Elliott, Olivia Riniker, Connor Boysen"
date: "2024-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on Water Quality Data

### Alexa Elliot, Olivia Riniker, Connor Boysen

## Introduction

This data set can asses the social and economic impacts that Covid-19 had on people in America. Using the frequency of symptoms of anxiety and depression over a seven day period.To asses these different variables we will explore these questions:

1.  What is the prevalence of anxiety and depression symptoms among different demographic groups in the United States during the COVID-19 pandemic?

2.  How have indicators of anxiety and depression changed over time during the survey period?

3.  Are there disparities in reported anxiety and depression symptoms based on factors such as age, gender, race/ethnicity, education level, or geographic location?

4.  What are the potential implications of the findings for public health interventions and mental wellness support during and after the pandemic?

5.  If a lock down were to ever happen again in what ways can we better address the emotional states of american citizens based on facts and statistics.

## Data Structure

the link for our data set is <https://data.cdc.gov/api/views/8pt5-q6wp/rows.csv?accessType=DOWNLOAD> . We obtained this data set from data.gov, it is a data set from the the U.S. Department of Health & Human Services which they obtained through an internet questionnaire.This data covers from April of 2020 to march or 2024 so we will be able to see the affects of the lock down as well as the change produced after the knockdown was disbanded.We are not include confidence interval(because we already have the seperate high and low), the quartile range(this column is missing many values), and time period label(because we have specific start and end dates.)

**pick variables to exclude that won't be helpful**

## cleaning data

First we load in the csv file for this data set.

```{r}
library(readr)
anxiety_depression <- read_csv("Indicators_of_Anxiety_or_Depression_Based_on_Reported_Frequency_of_Symptoms_During_Last_7_Days.csv")
View(anxiety_depression)

```

next we will start cleaning the data, starting with selecting the variables that we will be exploring in this data.

```{r}
# separating the data set by indicator 

# Filter data for depressive disorder
depressive_data <- anxiety_depression %>%
  filter(Indicator == "Symptoms of Depressive Disorder")

# Filter data for anxiety disorder
anxiety_data <- anxiety_depression %>%
  filter(Indicator == "Symptoms of Anxiety Disorder")

# Filter data for depressive and anxiety disorder
depressive_anxiety_data <- anxiety_depression %>%
  filter(Indicator == "Symptoms of Anxiety Disorder or Depressive Disorder")

```

hello