---
title: "Water Quality.rmd"
team members: "Alexa Elliott, Olivia Riniker, Connor Boysen"
date: "2024-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on Water Quality Data
###Alexa Elliot, Olivia Riniker, Connor Boysen

##Introduction
This data set can asses the social and economic impacts that Covid-19 had on people in America. Using the frequency of symptoms of anxiety and depression over a seven day period. 

1. What is the prevalence of anxiety and depression symptoms among different demographic groups in the United States during the COVID-19 pandemic?

2. How have indicators of anxiety and depression changed over time during the survey period?

3. Are there disparities in reported anxiety and depression symptoms based on factors such as age, gender, race/ethnicity, education level, or geographic location?

4. What are the potential implications of the findings for public health interventions and mental wellness support during and after the pandemic?

5. If a lock down were to ever happen again in what ways can we better address the emotional states of american citizens based on facts and statistics. 

```{r}
library(readr)
anxiety_depression <- read_csv("Indicators_of_Anxiety_or_Depression_Based_on_Reported_Frequency_of_Symptoms_During_Last_7_Days.csv")
View(anxiety_depression)

```

## cleaning data
```{r}

```
You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
